pirouette: Quantifying the importance of a tree prior in Bayesian phylogenetics

### 2019-03-19 article.tex
Updates from Overleaf

\begin{sidewaystable}
\centering
  \begin{tabular}{|@{}c|p{2.5cm}|p{9cm}|p{4.5cm}@{}|}
    \hline
    \centering
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \textbf{Symbol} & \textbf{Sub-argument} & \textbf{Description} & 
\textbf{Possible values} \\ 
    \hline
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{t_{p}}$ & tree prior & Macroevolutionary speciation model 
      & Yule, BD, CBS, CCP, CEP \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{c_{m}}$ & clock model & Clock for the DNA mutation rates 
      & strict, relaxed log-normal \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{s_{m}}$ & site model & Nucleotide substitution model 
      & JC, HKY, TN93, GTR \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{m_{r}}$ & mutation rate & Pace at which mutation occurs 
      & $m_{r} \in \mathbb{R}_{>0}$\\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{r_{s}}$ & root sequence & DNA sequence at the root of the tree 
      & any combination of a, c, g, t \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{m_{p}}$ & MRCA prior & Prior knowledge of a most recent common 
      ancestor & see documentation \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{m_{s}}$ & MCMC settings & Markov Chain Monte Carlo algorithm setup 
      & see documentation \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{m_{t}}$ & model type & Criterion to select an inference model 
      & Generative, Candidate \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{r_{i}}$ & run if & Condition under which an inference model is 
      used & Always, Best candidate\\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{e_{f}}$ & error function & Specifies how to measure the error 
      & nLTT, $|\gamma|$ \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{t_{m}}$ & twin tree prior & Sets the tree prior for the twin tree 
      & Yule, BD \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \hline
  \end{tabular}
  \caption{
    Most important parameter options. 
    BD = birth death (\cite{nee1994reconstructed}), 
    CBS = coalescent Bayesian Skyline (\cite{drummond2005bayesian}), 
    CCP = coalescent constant-population, 
    CEP = coalescent exponential-population,
    JC = Jukes-Cantor 1969 (\cite{jukes1969evolution}), 
    HKY = Hasegawa, Kishino and  Yano (\cite{hasegawa1985dating}), 
    TN93 = \cite{tamura1993estimation}, 
    GTR = Generalised time-reversible model (\cite{tavare1986some}).
  }
  \label{tab:options}
\bigskip

  \begin{tabular}{|@{}c|p{4cm}|p{9cm}|p{3cm}@{}|}
    \hline
    \centering
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \textbf{Symbol} & \textbf{Macro-argument} & \textbf{Description} 
      & \textbf{Includes} \\
    \hline
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{P}$ & Generative tree prior & Diversification model to generate 
      the 'true' tree & $\mathit{t_{p}}$ \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{A}$ & Alignment model & Model used to simulate the 
      alignment ($\mathit{c_{m}}$ must be strict) 
      & $\mathit{c_{m}}, \mathit{s_{m}}, \mathit{m_{r}}, \mathit{r_{s}}$ \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{G}$ & Generative model & The full setting to produce BEAST2 input 
      data & $\mathit{P}$, $\mathit{A}$ \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{I}$ & Inference model & Phylogenetic inference model to run BEAST2 
      & $\mathit{c_{m}}, \mathit{s_{m}}, \mathit{t_{p}}, 
      \mathit{m_{p}}, \mathit{m_{s}}$ \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{C}$ & Inference conditions & Conditions under which $\mathit{I}$ 
      is used in the inference & $\mathit{m_{t}}, \mathit{r_{i}}$\\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{X}$ & Experiment (candidate) 
      & Full setting for a Bayesian inference 
      & $\mathit{I},\mathit{C}$ \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{X'}$ & Experiment (selected) & As above & 
      $\mathit{I},\mathit{C}$, $Pr(D)$, $\omega$ \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{E}$ & Error measure parameters & Errors measurement setup & 
      $\mathit{e_{f}}$\\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{T}$ & Twinning parameters & Twin tree creation procedure & 
      $\mathit{t_{m}}$\\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \hline 
  \end{tabular}
  \caption{
    Definitions of terms and relative symbols used in the main text and in 
    Fig~\ref{fig:pipeline}. $Pr(D)$ denotes the evidence, $\omega$ denotes
    a model's weight 
  }
  \label{tab:definitions}
  \giovanni{
    Should we rename "Twinning Parameters" to "Twinning model"?
  }
  \richel{
    As the twinning parameters contain more than only the twin model's tree 
    prior, I suggest to keep it 'twinning parameters'
  }
  \giovanni{
    Should we rename "twin model" to "twin prior"?
  }
  \richel{
    I think that's a great idea! I would prefer the name 'twin tree prior' even
    better
  }
\end{sidewaystable}

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{workflow.png}
  \caption{
    \texttt{pirouette} pipeline. 
    The pipeline starts from a phylogeny (1a) simulated by generative prior 
$\mathit{P}$.
    The phylogeny is converted to an alignment (2a) using alignment model 
$\mathit{A}$. 
    The user supplies one or more experiments.
    For each candidate experiment $\mathit{X}$ 
    (a tuple of inference model $\mathit{I}$ and condition $\mathit{C}$),
    if its condition $\mathit{C}$ is 
    satisfied (which can depend on the alignment), 
    it becomes a selected experiment $\mathit{X'}$, 
    to be used in the next step (3a).
    The inference models of the selected experimenst use the alignment (2a) 
    to each create a Bayesian posterior of (parameter estimates and) 
    phylogenies (4a). 
    Each of the posteriors' trees is compared to the true phylogeny (1a) 
    using error measure $\mathit{E}$, 
    resulting in an error distribution (5a). 
    Optionally, a twin phylogeny (1b) can be generated from the original 
    phylogeny (1a) using twinning parameters $T$, where $T$ may depend on
    the experiments \richel{improve}.
    From a twin phylogeny, the same intermediate steps are taken,
    resulting in a twin error distribution (5b)
  }
  \label{fig:pipeline}
\end{figure}

@@

For all other steps we use the default setups. 
We can create the complete
\verb;pirouette; parameter set like this (which is the
same as listing~\ref{lst:create_pir_params}, but using the defaults):

@@

To answer ERQ 2 we used a evidence-based model selection procedure to obtain an 
error distribution.
However, it is unclear to what extent this error is due to the mismatch between 
the tree priors (from generative and inference models) or due to stochasticity. 
As mentioned in ERQ 1, \verb;pirouette; provides a tool to evaluate this, using 
the twinning process. We now apply it to the same pipeline followed in ERQ 2.
This leads to our fourth example research question:

>>>>

\begin{sidewaystable}
\centering
  \begin{tabular}{|@{}c|p{2.5cm}|p{9cm}|p{4.5cm}@{}|}
    \hline
    \centering
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \textbf{Symbol} & \textbf{Sub-argument} & \textbf{Description} & 
\textbf{Possible values} \\ 
    \hline
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{t_{p}}$ & tree prior & Macroevolutionary speciation model 
      & Yule, BD, CBS, CCP, CEP \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{c_{m}}$ & clock model & Clock for the DNA mutation rates 
      & strict, relaxed log-normal \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{s_{m}}$ & site model & Nucleotide substitution model 
      & JC, HKY, TN93, GTR \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{m_{r}}$ & mutation rate & Pace at which mutation occurs 
      & $m_{r} \in \mathbb{R}_{>0}$\\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{r_{s}}$ & root sequence & DNA sequence at the root of the tree 
      & any combination of a, c, g, t \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{m_{p}}$ & MRCA prior & Prior knowledge of a most recent common 
      ancestor & see documentation \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{m_{s}}$ & MCMC settings & Markov Chain Monte Carlo algorithm setup 
      & see documentation \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{m_{t}}$ & model type & Criterion to select an inference model 
      & Generative, Candidate \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{r_{i}}$ & run if & Condition under which an inference model is 
      used & Always, Best candidate\\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{e_{f}}$ & error function & Specifies how to measure the error 
      & nLTT, $|\gamma|$ \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{\tau_{p}}$ & twin tree prior & Sets the tree prior for the twin tree 
      & Yule, BD \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \hline
  \end{tabular}
  \caption{
    Most important parameter options. 
    BD = birth death (\cite{nee1994reconstructed}), 
    CBS = coalescent Bayesian Skyline (\cite{drummond2005bayesian}), 
    CCP = coalescent constant-population, 
    CEP = coalescent exponential-population,
    JC = Jukes-Cantor 1969 (\cite{jukes1969evolution}), 
    HKY = Hasegawa, Kishino and  Yano (\cite{hasegawa1985dating}), 
    TN93 = \cite{tamura1993estimation}, 
    GTR = Generalised time-reversible model (\cite{tavare1986some}).
  }
  \label{tab:options}
\bigskip

  \begin{tabular}{|@{}c|p{4cm}|p{9cm}|p{3cm}@{}|}
    \hline
    \centering
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \textbf{Symbol} & \textbf{Macro-argument} & \textbf{Description} 
      & \textbf{Includes} \\
    \hline
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{P}$ & Generative tree prior & Diversification model to generate 
      the 'true' tree & $\mathit{t_{p}}$ \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{A}$ & Alignment model & Model used to simulate the 
      alignment ($\mathit{c_{m}}$ must be strict) 
      & $\mathit{c_{m}}, \mathit{s_{m}}, \mathit{m_{r}}, \mathit{r_{s}}$ \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{G}$ & Generative model & The full setting to produce BEAST2 input 
      data & $\mathit{P}$, $\mathit{A}$ \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{I}$ & Inference model & Phylogenetic inference model to run BEAST2 
      & $\mathit{c_{m}}, \mathit{s_{m}}, \mathit{t_{p}}, 
      \mathit{m_{p}}, \mathit{m_{s}}$ \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{C}$ & Inference conditions & Conditions under which $\mathit{I}$ 
      is used in the inference & $\mathit{m_{t}}, \mathit{r_{i}}$\\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{X}$ & Experiment (candidate) 
      & Full setting for a Bayesian inference 
      & $\mathit{I},\mathit{C}$ \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{X'}$ & Experiment (selected) & As above & 
      $\mathit{I},\mathit{C}$, $Pr(D)$, $\omega$ \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{E}$ & Error measure parameters & Errors measurement setup & 
      $\mathit{e_{f}}$\\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{T}$ & Twinning parameters & Twin tree creation procedure & 
      $\mathit{\tau_{p}}$\\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \hline 
  \end{tabular}
  \caption{
    Definitions of terms and relative symbols used in the main text and in 
    Fig~\ref{fig:pipeline}. $Pr(D)$ denotes the evidence, $\omega$ denotes
    a model's weight 
  }
  \label{tab:definitions}
  \giovanni{
    Should we rename "Twinning Parameters" to "Twinning model"?
  }
  \richel{
    As the twinning parameters contain more than only the twin model's tree 
    prior, I suggest to keep it 'twinning parameters'
  }
  \giovanni{I don't completely get the $X'$ description. What are $Pr(D)$ and $\omega$? Is it something that we really need or it brings more complications than benefits?}
\end{sidewaystable}

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{workflow.png}
  \caption{
    \texttt{pirouette} pipeline. 
    The pipeline starts from a phylogeny (1a) simulated by generative prior 
$\mathit{P}$.
    The phylogeny is converted to an alignment (2a) using alignment model 
$\mathit{A}$. 
    The user supplies one or more experiments.
    For each candidate experiment $\mathit{X}$ 
    (a tuple of inference model $\mathit{I}$ and condition $\mathit{C}$),
    if its condition $\mathit{C}$ is 
    satisfied (which can depend on the alignment), 
    it becomes a selected experiment $\mathit{X'}$, 
    to be used in the next step (3a).
    The inference models of the selected experimenst use the alignment (2a) 
    to each create a Bayesian posterior of (parameter estimates and) 
    phylogenies (4a). 
    Each of the posteriors' trees is compared to the true phylogeny (1a) 
    using error measure $\mathit{E}$, 
    resulting in an error distribution (5a). 
    Optionally, a twin phylogeny (1b) can be generated from the original 
    phylogeny (1a) using twinning parameters $T$, where $T$ may depend on
    the experiments \richel{improve}.
    From a twin phylogeny, the same intermediate steps are taken,
    resulting in a twin error distribution (5b)
    \giovanni{Ok, that's my take on this: 1) all the capital letters should go in a ball (or a rhombus, as you already did for C) which stands in the middle of the arrows (as with C now); 2) The box in the middle should contain all the inference models $I_{i}$; 3) We need another C at the intersection of the arrows that point to T and A; 4) In between of the arrow 2a->4a there should be another rhombus (or circle) with an I (or I' or I*, according on how we want to call the selected inference model), which is pointed by the rhombus (of circle) containing C; 5) The first arrow should start from a rhombus (or circle) with P inside.}
  }
  \label{fig:pipeline}
\end{figure}

@@

For all the other steps we use the default setups. 
We can create the complete
\verb;pirouette; parameter set like this (which is the
same as listing~\ref{lst:create_pir_params}, but using the defaults):

@@

To answer ERQ 2 we used an evidence-based model selection procedure to obtain an 
error distribution.
However, it is unclear to what extent this error is due to the mismatch between 
the tree priors (from generative and inference models) or due to stochasticity. 
As mentioned in ERQ 1, \verb;pirouette; provides a tool to evaluate this, using 
the twinning process. We now apply it to the same pipeline followed in ERQ 2.
This leads to our fourth example research question:
